Online learning methods for financial time series forecasting

### 2017-12-13 slides/defense-slides.tex
Contribucion esquema y mejor motivacion

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: High Frequency Trading
\item Cointegration and the VECM model
\item Challenges and Hypothesis
\item Contributions
\item Limitations and Applicability
\item Conclusions and Future Research
\end{itemize}
\end{frame}

@@

@@

@@

\begin{frame}
\frametitle{Cointegration}
Cointegration reflects the idea of that some set of non-stationary variables cannot wander too far from each other.
\begin{figure}
\includegraphics[width=0.8\paperwidth]{img/motivation}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Cointegration definition}
%The idea of cointegration is that there is an I(1) process, underlying two or more processes.
\begin{block}{Cointegration}
The relationship between non-stationary time series can be modelled if a {\bf stationary linear combination} is shown to exist.
When this happens it is said they have a {\bf long-run
equilibrium} relationship and the variables are {\color{red} cointegrated}.
\end{block}
\end{frame}

@@

The Vector Error correction model (VECM) is used to model cointegrated time series:

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} {\boldsymbol\phi}_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\boldsymbol{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
\noindent where $\mathbf{y}_t = [y^1_t,\dots,y^l_t]^\intercal$, $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$. The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
\end{frame}

\begin{frame}
\frametitle{$\boldsymbol \Omega$ matrix properties}
The matrix $\boldsymbol \Omega$ has the following properties:
\begin{itemize}
\item If ${\color{blue}\boldsymbol \Omega = 0}$ there is no cointegration
\item If ${\color{blue}rank(\boldsymbol\Omega)=l}$ i.e full rank, then the time series are not
I(1) but stationary
\item If ${\color{blue}rank(\boldsymbol\Omega)=r,\quad 0 < r < l}$ then, there is cointegration.
%where $\boldsymbol\alpha$ and $\boldsymbol\beta$ are $(l \times r)$
%matrices and $rank(\boldsymbol\alpha)=rank(\boldsymbol\beta)=r$.
\end{itemize}
\end{frame}

@@

\section{Contributions}
\begin{frame}
\frametitle{Adaptive VECM}
\hspace*{-9mm}
\includegraphics[width=1.15\textwidth]{img/slidingwindow}
\end{frame}

>>>>

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: High Frequency Trading and Cointegration
\item Challenges and Hypothesis
\item Contributions
\item Limitations and Applicability
\item Conclusions and Future Research
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Motivation: HFT}
\begin{block}{High Frequency Trading}
High Frequency Trading (HFT) is a technology that implements different trading strategies to buy or sell an stock using sophisticated infrastructure and high speed algorithms.
\end{block}

Today, the volume of data has rapidly increased:

\begin{table}
\begin{tabular}{ c | c }
Period & Data precision \\
\hline \hline
1993-2003 & seconds \\
2003-2015 & milliseconds \\
27th July 2015-onwards & microseconds \\
24th October 2016-onwards & nanoseconds
\end{tabular}
\caption{High Frequency data precision}
\end{table}

@@

\begin{frame}
\frametitle{Motivation: HFT}
\begin{alertblock}{Arbitrage}
The practice of taking advantage of market inefficiency or discrepancies in the market price.
\end{alertblock}
\begin{columns}

\begin{column}{0.4\textwidth}
\begin{exampleblock}{Advantages}
\begin{itemize}
\item Increases liquidity
\item Allows more efficient price formation
\item Reduce the difference between buyers and sellers prices.
\end{itemize}
\end{exampleblock}
\end{column}
\begin{column}{0.4\textwidth}
\begin{exampleblock}{Risks}
\begin{itemize}
\item Removal of human decision making could lead to failures in untested or incorrect strategies
\end{itemize}
\end{exampleblock}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Motivation: Cointegration}
Cointegration reflects the idea of that some set of non-stationary variables cannot wander too far from each other.
\begin{figure}
\includegraphics[width=0.8\paperwidth]{img/motivation}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Contribution}
This thesis contribution has several components:
\begin{figure}
\includegraphics[width=\textwidth]{img/contribution}
\end{figure}
\end{frame}

@@

\begin{frame}
\frametitle{Cointegration definition}
%The idea of cointegration is that there is an I(1) process, underlying two or more processes.
\begin{block}{Cointegration}
The relationship between non-stationary time series can be modelled if a {\bf stationary linear combination} is shown to exist.
When this happens it is said they have a {\bf long-run
equilibrium} relationship and the variables are {\color{red} {\bf cointegrated}}.
\end{block}
\begin{figure}
\includegraphics[width=0.8\paperwidth, height=0.4\paperheight]{img/motivation}
\end{figure}
\end{frame}

@@

The Vector Error correction model (VECM), considering $p$ lags, is used to model cointegrated time series:

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} {\boldsymbol\phi}_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\boldsymbol{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
\noindent where 
\begin{itemize}
\item $\mathbf{y}_t = [y^1_t,\dots,y^l_t]^\intercal$,
\item the errors $\mathbf{e}_t$ are assumed to follow i.i.d l-dimensional multivariate normal distribution $\mathcal{N}({\bf O,\Sigma})$, 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Vector Error correction}
\tikzstyle{na} = [baseline=-.5ex]

The Vector Error correction model (VECM), considering $p$ lags, is used to model cointegrated time series:

{\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} {\boldsymbol\phi}_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\boldsymbol{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
}
\noindent where 
\begin{itemize}
\item $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$. The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{$\boldsymbol \Omega$ matrix properties}

Since $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$ we can also say that:

$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \boldsymbol \beta^\intercal \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)}$$

The matrix $\boldsymbol \Omega$ has the following properties:
\begin{itemize}
\item If ${\color{blue}rank(\boldsymbol \Omega) = 0}$ there is no cointegration
\item If ${\color{blue}rank(\boldsymbol\Omega)=l}$ i.e full rank, $\boldsymbol\Omega$ is non-singular or invertible, therefore we can express $\mathbf{y}$ as follows:
$$\mathbf{y} = \boldsymbol \Omega ^{-1} \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, ,$$
then the time series are not I(1) but stationary.
\item If ${\color{blue}rank(\boldsymbol\Omega)=r,\quad 0 < r < l}$ then, r cointegration vectors exist such that: 
$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, .$$
%where $\boldsymbol\alpha$ and $\boldsymbol\beta$ are $(l \times r)$
%matrices and $rank(\boldsymbol\alpha)=rank(\boldsymbol\beta)=r$.
\end{itemize}
\end{frame}

@@

\section{Contributions}

\begin{frame}
\frametitle{Adaptive VECM}
\hspace*{-9mm}
\includegraphics[width=1.15\textwidth]{img/slidingwindow}
\end{frame}
